String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-01 2023-1-19/sections/algorithms.tex
algorithm

After encoding counting operator and length constraint to CEFAs, there are only automaton memberships $x\in \aut$ and linear integer arithmetic $t_1 \ \op \ t_2$ where $\aut$ is CEFA and $t_1, t_2$ are integer terms. Let $\psi$ denotes the conjunction of all LIA formulas $t_1\ \op\ t_2$ and $\Lambda_x = \{\aut_x^1, \aut_x^2,\cdots \aut_x^n\}$ denotes the set of CEFAs in the automaton membership $x\in \aut$. The emptiness checking problem of $\Lambda_x$ under linear integer arithmetic $\psi$ is to check whether there exists a word $w$ such that $w\in \aut_x^1, w\in \aut_x^2\cdots w\in \aut_x^n$ and $\psi[\myvec{V}/\myvec{r}]$ holds, where $\myvec{r}$ are all registers and $\myvec{V}$ are corresponding register values in automata of $\Lambda_x$ after running $w$. The emptiness problem is \emph{abbreviated as $SAT_{CEFA}[LIA]$ problem}. \newline

@@

The main idea of our framework is to simplify the input CEFAs firstly and under-approximate the solution of the \emptinessprob{} by adding length bounds on string variables. If the under-approximation is satisfiable, the CEFAs is not empty under $\psi$. Otherwise, we check the Parikh image of the simplified CEFAs. If the Parikh image is satisfiable, the CEFAs under linear integer arithmetic $\psi$ are not empty. Or else, they are empty. The framework is \emph{sound} and \emph{complete}.\newline
The framework is shown in Algorithm \ref{alg:emptiness}, we simplify the input CEFAs at line 1 because they may have many transitions and registers. The main idea of the simplification is treating alphabet as unary. The symbolic updates of the CEFAs are maintained, which are then treated as labels of transitions in the minimization. After the simplification, we try to find a solution by under-approximation at line 3 and line 4. An off-the-shelf SMT solver gives the result of LIA $\varphi_{under}\wedge \psi$. If the LIA is satisfiable, we return $true$ at line 5 because it implies that the CEFAs under the LIA $\psi$ are not empty. Otherwise, we compute the Parikh images of the simplified CEFAs and check if the Parikh images are satisfiable in conjunction with $\psi$ at line 12. The Parikh image checking is complete so that satisfiability implies non-emptiness directly. \newline
The incremental under-approximation program is executed many times until the $MaxBound$ is reached. $MaxBound$ is an empirical bound set to $15$ in our experiment because it brings exponential growth of searching space. The details of simplification are shown in Algorithm \ref{alg:simplify} and under-approximation are shown in Algorithm \ref{alg:underApprox}.\newline

@@

The purpose of simplification is to remove duplicated transitions and registers. In the emptiness checking Algorithm \ref{alg:emptiness}, the non-emptiness is equal to the reachability of CEFAs. Thus, the symbolic updates on transitions are meritorious because it influences registers values, while the characters are not. The alphabet of the CEFA can be seen as unary (i.e., the alphabet is $\{a\}$) and the vectors as labels. The special vector $\myvec{0}_n$ with only zero elements is seen as $\epsilon$ label. Namely, given a CEFA $\aut = (Q, \Sigma, \delta, q_I, F, R, \theta)$, we obtain a symbolic form  $\aut_{sym} = (Q, \Sigma', \delta', q_I, F, R, \theta)$ where

@@

 For the symbolic form $\aut_{sym}$, we treat $q\xrightarrow[\myvec{0}_n]{a}q'$ as epsilon transition and compute the epsilon-closure based on it. Likewise, we treat $q\xrightarrow[\myvec{v}]{a}q'$ as a transition with label $(a, \myvec{v})$ and determine the automaton based on the label. Two accepting states are equivalent if they have the same accepting condition, and two states are equivalent if they reach equivalent states for each label. We simplify the automaton by merging these equivalent states. Briefly, we directly apply algorithms of epsilon-closure, determination, and simplification in \cite{aut_hopcraft}.

>>>>

After encoding counting operator and length constraint to CEFAs, there are only automaton memberships $x\in \aut$ and LIA formula $t_1 \ \op \ t_2$ where $\aut$ is CEFA and $t_1, t_2$ are integer terms. Let $\psi$ denotes the conjunction of all LIA formulae $t_1\ \op\ t_2$ and $\Lambda_x = \{\aut_x^1, \aut_x^2,\cdots \aut_x^n\}$ denotes the set of CEFAs in the automaton membership $x\in \aut$. The emptiness checking problem of $\Lambda_x$ under $\psi$ is the problem that checks whether there exists a word $w$ such that $w\in \aut_x^1, w\in \aut_x^2\cdots w\in \aut_x^n$ and $\psi[\myvec{V}/\myvec{r}]$ holds, where $\myvec{r}$ are all registers and $\myvec{V}$ are corresponding register values in automata of $\Lambda_x$ after running $w$. The emptiness problem is \emph{abbreviated as $SAT_{CEFA}[LIA]$ problem}. \newline

@@

The main idea of our framework is to simplify the input CEFAs firstly and under-approximate the solution of the \emptinessprob{} by adding length bounds on string variables. If the under-approximation is satisfiable, the CEFAs are not empty under $\psi$. Otherwise, we check the Parikh image of the simplified CEFAs. If the Parikh image is satisfiable, the CEFAs are not empty under $\psi$. Or else, they are empty. The framework is \emph{sound} and \emph{complete}.\newline
The framework is shown in Algorithm \ref{alg:emptiness}, we simplify the input CEFAs at line 1 because they may have too many transitions and registers to solve. The main idea of the simplification is treating the alphabet as unary and the symbolic updates as characters of transitions in the minimization. After the simplification, we try to find a solution by under-approximation at line 3 and line 4. An off-the-shelf SMT solver gives the result of LIA $\varphi_{under}\wedge \psi$. If the LIA is satisfiable, we return $true$ at line 5 because the CEFAs under the LIA $\psi$ are not empty. Otherwise, we compute the Parikh images of the simplified CEFAs and check if the Parikh images are satisfiable in conjunction with $\psi$ at line 12. The Parikh image checking is complete so that satisfiability implies non-emptiness directly. \newline
The incremental under-approximation program is executed many times until the $MaxBound$ is reached. $MaxBound$ is an empirical bound set to $15$ in our experiment because it brings exponential growth of searching space. The details of simplification and under-approximation are shown in Algorithm \ref{alg:simplify} and the details of  are shown in Algorithm \ref{alg:underApprox}.\newline

@@

The purpose of simplification is to remove duplicated transitions and registers. In the emptiness checking Algorithm \ref{alg:emptiness}, the non-emptiness is equal to the reachability of CEFAs. Thus, the symbolic updates on transitions are meritorious because it influences registers values, while the characters are not. The alphabet of the CEFA can be seen as unary (i.e., the alphabet is $\{a\}$) and the vectors as characters. The special vector $\myvec{0}_n$ with only zero elements is seen as $\epsilon$ character. Namely, given a CEFA $\aut = (Q, \Sigma, \delta, q_I, F, R, \theta)$, we obtain a symbolic form  $\aut_{sym} = (Q, \Sigma', \delta', q_I, F, R, \theta)$ where

@@

 For the symbolic form $\aut_{sym}$, we treat $q\xrightarrow[\myvec{0}_n]{a}q'$ as epsilon transition and compute the epsilon-closure based on it. Likewise, we treat $q\xrightarrow[\myvec{v}]{a}q'$ as a transition with character $(a, \myvec{v})$ and determine the automaton based on it. Two accepting states are equivalent if they have the same accepting condition, and two states are equivalent if they reach equivalent states for each character. We simplify the automaton by merging these equivalent states. Briefly, we directly apply algorithms of epsilon-closure, determination, and simplification in \cite{aut_hopcraft}.
